---
title: "Analyze MS-CETSA melt curves"
author: "Dai Lingyun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{analysis of CETSA melt curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction  
This script is designed to process CETSA melt curve result from TMT10 labelling and quantification scheme.  
The main tasks include:  
1. Read in some data (Search result txt.file from Proteome Discoverer, PD) into R/Rstudio  
2. Clean up data, mainly to remove the entries without quantitative information  
3. Normalize data based on the principle that the overall median of relative protein fold changes across the heating temperature range should follow a sigmoidal trend  
4. Try to fit each entry of data into a temperature-response relationship and capture the fitting parameters  
5. Visualize the data using customizable plot layout, for either all the data or the subsets  

--------------------------------------------------------------------------------------------

## Prerequisites
* R version > 3.3.0 
* Rstudio version > 0.99, preferably > 1.0
* Dependent packages (will be handled during the package installiation process automatically):  
"readr", "drc", "plyr", "tibble", "tidyr", "grid", "gtools", "gridExtra", "scales", "RColorBrewer", "dplyr", "ggplot2"  
However, if complained, use `install.packges()` command to manually install a paticular package  

--------------------------------------------------------------------------------------------


## Proceessing
1. For better organization of data analysisk, it is __highly recommended__ to use the [Project management feature](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects) in Rstudio.   
For each project, you are suggested to create a brand new working folder as your local working directory.  
Save the PD derived data files to this working directory folder.  
At the right up corner of Rstudio, click Project -> New Project -> Existing Directory -> browse and choose the designated working directory folder.  
By establishing a specfic project for each experiment, all the data content and execution history will be saved in the project for future analysis even when R session is terminated.  

2. Activate `mineCETSA` package by library it.  
```{r, message=TRUE}
library("mineCETSA")
```

3. Read in data from result files, `ms_rawread()` is designed for read in one or many files in a vector of file names.  
By default, `PD21=TRUE` assume the data is searched using PD2.1; otherwise, specify `PD21=FALSE` if data is originated from PD2.0  
Refer to "Data searching using PD2.1" for how-to.  
Note that for simplicity, always use channel __126__ as control channel for data search in PD, regardless of the labelling arrangement.    
You need to specify a `temp` (short for temperature) vector that was the heat challenge the samples/cells were exposed to in CETSA treatment, corresponding to the labelling done from 126 up to 131, of which the function default is `temp=c(37,40,43,46,49,52,55,58,61,64)`.  
```{r, eval=FALSE}
LY <- ms_rawread(c("LY_K562_Ctrl_rep1_PD21_Proteins copy.txt","LY_K562_Ctrl_rep2_PD21_Proteins copy.txt","LY_K562_Treatment_rep1_PD21_Proteins copy.txt","LY_K562_Treatment_rep2_PD21_Proteins copy.txt"))
```

```{r, echo=FALSE, message=TRUE}
LY <- ms_rawread(c(system.file("extdata", "LY_K562_Ctrl_rep1_PD21_Proteins copy.txt", package="mineCETSA"), system.file("extdata", "LY_K562_Ctrl_rep2_PD21_Proteins copy.txt", package="mineCETSA"), system.file("extdata", "LY_K562_Treatment_rep1_PD21_Proteins copy.txt", package="mineCETSA"), system.file("extdata", "LY_K562_Treatment_rep2_PD21_Proteins copy.txt", package="mineCETSA")))
```

4. Customize the condition names in dataset using `ms_conditionrename()`.  
When read in multiple files, a number suffix is added to the condition of each data file in sequence   
A format of _"Experimental conditon.Replicate number"_ is suggested. More specifically, the following three formats are supported: 1) condition names followed by dot and number (eg, DMSO.1, 37C.2); 2) condition names followed by dot and rep and number (eg, DMSO.rep2, 52C.rep1); 3) condition names followed by dot and letter r and number (eg, Ctrl.r1, dTITTR37C.r2).  
__It is recommended that you should try to omit using dot in the experimental condition names, you can segregate your long condition name with underline if necessary.__  
The argument `incondition` corresponses to current condition naming, whereas `outcondition` correponses to new condition naming, both order and length should match exactly.  
```{r}
LY1 <- ms_conditionrename(LY, incondition=c("Treatmentrep1.3", "Treatmentrep2.4"), outcondition=c("Treatment.1","Treatment.2"))
```

5. Remove not-quantified and Keratin proteins by default using `ms_clean()`.  
By default the orphan proteins that appear only in one of the datasets is not removed, however you can specify `remsinglecondprot=TRUE`, these orphan proteins will be excluded from downstream analysis.  
```{r}
LY_cleaned <- ms_clean(LY1)
```

6.1. Apply systematic scaling to the dataset using `ms_scaling()`.  
The function achieves: The medians of soluble protein ratio is fitted into a sigmoidal function of the temperature range   
Output in working directory:  
* Scaling factors used is written into a local file   
* Medians of pre- and post- data normalization soluble protein ratio is plotted  
* The distribution of pre- and post- normalization soluble protein ratio is plotted  
```{r}
LY_scaled <- ms_scaling(LY_cleaned)
```

6.2. The alternative for systematic scaling is to force scale the median of data to be same as the median from a subset of data by `ms_forcescaling()`. However, make sure you have enough good reason to perform forcescaling.  
```{r}
LY_fs_all <- ms_forcescaling(LY_cleaned, refcondition="all")
LY_fs <- ms_forcescaling(LY_cleaned, refcondition=c("Ctrl.1", "Ctrl.2"))
```

7. Try to perform temperature-response function fitting to generate fitting parameters for each entry/curve using `ms_fitting()`.  
The returned data contains the input scaled data plus fitting parameters (including `Tm`, `R2`, `Slope`, `RSE`(Residual standard error)).  
A copy of the fitted dataset would be saved in current working directory.  
```{r, eval=FALSE}
LY_fitted <- ms_fitting(LY_scaled)
```

8.1. Plot out melt curves using `ms_ggplotting()`.  
This function is previously designed for single replicated experiment, mainly about the simplicity and generality.   
The resulting plots were sorted based on the difference of AUCs of the same protein under different experimental conditions, i.e., the proteins showing most significant "shift"/change would be ranked at the top of the file. However, it has been observed that noisy readings tend to mess up the rankings.  
```{r, eval=FALSE}
ms_ggplotting(LY_scaled)
```

8.2. Plot out replicated melt curves using `ms_ggplotting_rep()`.  
This function mainly cater for the (technical or biological) replicated runs of CETSA melt curves, the ranking system not only considers the nominal shifts between different experimental conditions, but also the internal reproducibility among the replicate runs.  
The arguments and its default value includes:  
* `nread=10`, how many readings per curve (from lowest dose) were used for plotting, default is 10  
* `levelvector`, a vector of conditions to be specified, preferably starting with Ctrl sets  
* `remsinglecondprot=TRUE`, whether orphan proteins to be plotted, default to exclude them from plotting  
* `nreplicate=2`, how many replicates there are for each condition, default is duplicate thus 2, change the value accordingly, this is relavent to automatic coloring scheme, up to 4 is possible for now  
* `layout=c(5,5)`, the default multiplot layout is 5 times 5 in each page  
* `presetcolor=TRUE`, whether automatically apply the preset coloring scheme for each condition, otherwise, provide a vector for example `colorpanel=c("black", "red", "blue", "orange")` to customize the colors  
* `dotconnect=FALSE`, by default to fit the best melt curve, however, by specifing `dotconnect=TRUE` you could skip the curve fitting  
```{r, eval=FALSE}
ms_ggplotting_rep(LY_scaled, levelvector=c("Ctrl.1", "Ctrl.2", "Treatment.1","Treatment.2"))
```

9. Export the scaled data (virtually any data in the current working space) using `ms_filewrite()` to a local txt file.
```{r, eval=FALSE}
ms_filewrite(LY_scaled, "CETSA_data_scaled.txt")
ms_filewrite(LY_fitted, "CETSA_data_with_fitting_parameters.txt")
```

10. Read in the exported data (virtually any data in the current working space) using `ms_fileread()` from a saved txt file.
```{r, eval=FALSE}
MyCETSAdata <- ms_fileread("./subfolder/myCETSAdata_to_readin.txt")
# . indicate the current working directory
# Here the subfolder refer to the first level folder name under current working directory in which the data file locates
# Remember that typically now you are still working in the main working directory
```

11. Focus on a list of target proteins to study using `ms_subsetting()`.  
You need to make an excel file with a list of Uniprot IDs you want to study (for example, P00000, P12345-6) under the header “__id__”, then save as txt file in working directory, say under the name “hit_list.txt”. The alternative for a short list, you can simply provide a vector of Uniprot IDs.  
The arguments and its default value includes:  
* `isfile=TRUE`, whether the provided hitidlist is in a txt. file, default is set to TRUE
* `allisoform=TRUE`, whethre to retrive other isoforms of the same parental Uniprot ID, default is set to TRUE
* `revsel=FALSE`, short for reverse selection, when set to TRUE, the output is the dataset after removing the targets, default is set to FALSE
```{r, eval=FALSE}
CETSAdata_subset <- ms_subsetting(LY_fitted[[1]], hitidlist="hit_list.txt", isfile=TRUE)
CETSAdata_subset <- ms_subsetting(LY_fitted[[1]], hitidlist=c("P00000", "P12345-6"), isfile=FALSE)
```

